Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-04 paper/paper.qmd
committing changes to paper.qmd

---
title: "Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory"
subtitle: A Bayesian Analysis of Polling Data from Key Battleground States"
author: 
  - Jimin Lee
  - Sarah Ding
  - Xiyan Chen
thanks: "Code and data are available at: [https://github.com/jamiejiminlee/2024_US_Elections.git](https://github.com/jamiejiminlee/2024_US_Elections.git)."
date: today
date-format: long
abstract: "This paper presents a predictive model for the 2024 United States Presidential Election, focusing on critical battleground states. Utilizing a Bayesian approach with state-level polling data, we estimate the winning probabilities for candidates Donald Trump and Kamala Harris. Our findings reveal that Harris is projected to win in six out of seven key battleground states, including Michigan, Nevada, and North Carolina, with predicted winning percentages surpassing 50%. Given this significant advantage, we predict Kamala Harris to be the likely winner of the election. This analysis underscores the importance of battleground states in shaping electoral outcomes and provides valuable insights into voter sentiment as Election Day approaches."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction {#sec-intro}
The 2024 United States Presidential Election is shaping up to be a critical event, particularly as battleground states emerge as pivotal areas influencing the electoral outcome. As candidates Donald Trump and Kamala Harris intensify their campaigns, understanding voter preferences in these key regions becomes essential for predicting the election results. This paper utilizes state-level polling data from Project 538 [@FiveThirtyEight] and the statistical programming language R [@citeR] to build a predictive model that estimates the likelihood of victory for each candidate in several key battleground states, including Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin.

The primary estimand in this study is the probability that either Donald Trump or Kamala Harris will win each battleground state, based on aggregated polling averages. Our model employs a Bayesian hierarchical approach with natural splines to capture the nuances of voter sentiment over time, allowing for the estimation of winning probabilities across these crucial states. By comparing the predicted winning probabilities for both candidates, we can ascertain the likely winner based on the prevailing voter support patterns.

Our analysis reveals that Kamala Harris is projected to win in six out of the seven battleground states examined, indicating a favorable voter sentiment in states such as Michigan and Nevada. Conversely, Donald Trump shows a competitive stance in Arizona, where he is also predicted to secure a slight edge. The calculation of the predicted winner is determined by comparing the estimated probabilities for each candidate, which underscores Harris's overall advantage in the election.

Understanding these predictions is vital as it offers insights into the dynamics of voter behavior and the potential implications for election outcomes. This information can help political analysts, campaign strategists, and the public anticipate how these battleground states may influence the overall results. Accurate predictions not only inform campaign strategies but also highlight regions where voter sentiment may shift in the lead-up to Election Day.

The remainder of this paper is structured as follows. In @sec-data, we describe the data and variables used in the analysis. @sec-model outlines the model setup and estimation strategy. @sec-results presents the results, including the predicted election outcomes and visualizations. Finally, @sec-discussion discusses the implications and limitations of our findings.

# Data {#sec-data}
## Overview
The dataset of the 2024 Presidential Election cycle is obtained from Project 538 [@FiveThirtyEight]. The dataset is periodically updated throughout the presidential election campaign to reflect current and most up-to-date polling results. It consists of polling information from various pollsters such as the New York Times, ActiVote, Morning Consult, etc. For every poll conducted, relevant variables such as state, start_date, end_date, transparency_score, candidate_name, and percentage of support (pct) for the appropriate candidates are included. The dataset compiles data from all major pollsters in the US and provides us with an understanding of the current state of the presidential election.

The present analysis focuses on Donald Trump and Kamala Harris, the two leading candidates in the current presidential election race. For all the following variables mentioned, we collected observations for both of the candidates. We collect observations from the dataset where the ‘candidate_name’ is Trump and Harris. We are interested in the support for both candidates over time, for each state, therefore, we collect observations of ‘end_date’, and ‘state’, as well as the electorate support for each candidate is denoted as ‘pct’. Furthermore, since there are many pollsters present in the dataset, we extracted pollsters with high-quality data that is reliable, which is denoted as having a high ‘numeric_grade’ and ‘transparency_score’, details of data cleaning are further explained in [Appendix -@sec-data-details].

## Measurement
In the current presidential election cycle,  Americans’ opinions on voting for their preferred candidate are transformed into data points through a series of steps that involve surveying, data processing, and structuring responses. Pollsters then store these responses in structured databases for analysis. In the dataset obtained from Project 538 [@FiveThirtyEight], each entry in the dataset corresponds to a poll conducted by each pollster, capturing the percentage of respondents who express support for each candidate. Each pollster has different approaches to sampling respondents, recruiting respondents, as well as questionnaire types, which affects each pollster’s reliability and data quality.

Pollsters begin by selecting a sample from the registered voter population, often using voter files with demographic and contact information. They then apply methods of stratified random sampling to ensure the sample reflects the electorate’s diversity in terms of age, gender, race, education, voter behavior, etc.

Then, pollsters reach out to voters through various channels, such as phone calls, emails, text messages, or online panels. These are called recruiting methods and usually, each of them has its trade-offs. If the respondents are connected through any of these channels, they are asked a series of questions either by a live interviewer or a preset of questions.

The types of questions often include:

- “If the election were held today, would you vote for Candidate A or Candidate B?” 
- “How strongly do you support Candidate?” (e.g., on a scale of 1 to 10, or strongly, somewhat, undecided)
- “Do you support Policy A proposed by Candidate A or Policy B proposed by Candidate B?”

Responses are recorded digitally, either by survey software or directly into a database by interviewers, capturing both the answers and relevant demographic data. Each response becomes a data point in the database. For example, if a respondent prefers Candidate A, their answer may be recorded as a binary variable (e.g., “1” for Candidate A, “0” for Candidate B). Along with candidate preference, pollsters capture demographic and behavioral details, such as age, gender, race, education, income, voter history, and party registration.

Each respondent’s answers and demographic details are grouped and saved in a structured database. Pollsters then clean the data by checking for inconsistencies, removing incomplete responses, and addressing non-responses. Pollsters also perform weighting to ensure the representativeness of data collected, details of weighting are further explained in @sec-pollster-methodology.

Once cleaned and weighted, data points are aggregated to determine overall candidate support, often producing metrics such as the percentage of respondents supporting each candidate by state or other specific demographics. Each pollster would have a database with all the weighted and aggregated data, and Project 538 presents a compilation of the databases obtained from various pollsters.

## Variables
The collected dataset contains several key variables relevant to the analysis:

@@

To ensure the quality of the analysis, only those polls that meet specific thresholds were included - for more details, refer to [Appendix -@sec-data-details].

@@

# Display vertical summary statistics in a kable table
kable(summary_statistics_long) %>%
  kable_styling(full_width = FALSE, position = "center")

```
For instance, @tbl-outcomevar reveals a total of 107 polls conducted, calculated as the total count of rows in our cleaned dataset. ‘Total polls’ reflects the total number of unique poll entries after filtering for only battleground states, polls with sufficient data quality, and polls that report support percentages for both Trump and Harris. The 107 polls conducted display an average support level of 47.801% for Trump and 47.999% for Harris. This suggests that based on the dataset, Trump has a leading advantage in winning the presidential election race by having a slighty higher average level of support than Harris. The highest percentage of support for Trump in the 107 polls was 51.700% while the lowest recorded support was 43%. For Harris, the highest percentage of support also reached 51.200%, with a minimum of 43.600%. These statistics illustrate the competitive landscape of the presidential election race, without a clear leading candidate in the current race.

@@

#### State
The state ('state') variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of Observations for Each State in Polling Data. Illustrates the number of polling observations across key battleground states in the lead-up to the 2024 United States Presidential Election. Notably, Pennsylvania has the highest count of observations, suggesting a more robust polling presence, while Nevada shows fewer observations. "

@@

#### End Date {#sec-end-date}
End date ('end_date') represents the date the poll ended, essentially the date the poll stopped collecting data from electorates. In the context of our paper, the end date of each poll serves as one of the predictive variables and is used to assess changes in voter support over time. Including end date as a variable in our model can help us identify patterns in voter support as Election Day approaches, potentially reflecting the impact of campaign events, debates, or other political developments. @fig-end-date-distribution displays the distribution of polling end dates from July 2024 to October 2024.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Distribution of Polling End Dates. This histogram displays the frequency of polling end dates for key battleground states leading up to the 2024 United States Presidential Election, highlighting peaks in polling activity around late September."

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-numgrade-distribution
#| fig-cap: "Density plot illustrating the distribution of numeric grades assigned to pollsters within the polling data as the 2024 United States Presidential Election approaches. The graph highlights the frequency of numeric grades, indicating key concentrations that suggest variability in pollster quality. "

@@

#### Transparency Score
Transparency Score ('transparency_score') represents the degree of methodological transparency provided by each pollster, rated on a scale from 0 to 10 by [@FiveThirtyEight]. This score reflects the extent to which pollsters disclose details about their data collection and analysis methods, such as sample size, weighting, and margin of error. In the context of our paper, transparency score is an essential variable for evaluating the reliability of polling data, as higher transparency scores generally indicate a greater level of methodological rigor and openness. @fig-transparency-score-distribution displays the distribution of transparency scores across the polling data for key battleground states. The plot shows that transparency scores cluster heavily around ratings of  7 and 9, suggesting that a majority of polls in our dataset come from sources that maintain a relatively high level of transparency. There are smaller peaks at lower scores (5 and 10), indicating that while some pollsters offer limited transparency, a few polls also achieve perfect transparency scores.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram displaying the distribution of transparency scores among pollsters in the polling data leading up to the 2024 United States Presidential Election. The plot indicates the frequency of various transparency scores, highlighting a concentration of scores around 7 and 9, which suggests a predominance of high transparency among the assessed pollsters. "

@@

#### Pollster
Pollster ('pollster') represents the various organizations responsible for conducting polls across key battleground states in the lead-up to the 2024 United States Presidential Election. In the context of our paper, the diversity and frequency of polling by different pollsters are essential for understanding the reliability and variance in the data. Each pollster may use different methodologies, sample populations, and data collection techniques, which can affect the results and introduce variability in polling outcomes.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-pollster-distribution
#| fig-cap: "Horizontal bar chart displaying the distribution of polls conducted by various pollsters in key battleground states for the 2024 United States Presidential Election. The chart highlights the frequency of polling activities across different organizations, with Marist and Emerson standing out as the most prolific pollsters."

@@

#### Start Date
Start date ('start_date') represents the date when each poll began collecting responses. In the context of our paper, start date is a key variable as it allows us to focus on polling data collected after July 21, 2024—the date Kamala Harris officially announced her presidential campaign. By filtering the dataset to include only polls with a start date after this announcement, we ensure that each poll reflects voter sentiment with Harris as an active candidate, enabling a more accurate comparison between her and Donald Trump. @fig-start-date-distribution displays the distribution of polling start dates from late July to mid-October 2024, showing when polling efforts began across key battleground states. The histogram reveals peaks in polling activity, particularly in mid-August, early September, and late September, indicating periods of heightened interest or campaign events that likely prompted increased polling.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-start-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling start dates for key battleground states leading up to the 2024 United States Presidential Election. The chart highlights the frequency of polling activities initiated over time, with a notable peak on September 14, indicating a surge in polling efforts as the election approached."

@@

## Model Justification

The Bayesian Beta Regression model with a logit link function is the most suitable model for our prediction of the 2024 U.S. Presidential Election because it aligns closely with both the nature of the data and the goals of our paper, which include understanding voter support trends over time and accounting for regional differences across battleground states.

Our primary outcome variable is the percentage of support for each candidate in 7 key battleground states, scaled to a proportion between 0 and 1. Support percentages are naturally bounded between 0% and 100%, and this bounded nature poses challenges for traditional linear regression, which assumes an unbounded continuous outcome. The Beta distribution is ideal for modeling proportion data because it is inherently defined on the interval (0, 1). This allows us to model the proportions of support without the risk of generating unrealistic values outside this range. Unlike a linear model, which could predict support percentages exceeding 100% or falling below 0%, the Beta distribution respects the boundaries of our data. The logit link function further ensures that the model predicts in terms of log odds, which is suitable for interpreting support as probabilities—an essential consideration given our goal of forecasting potential victory in each state.

One of our key goals is to examine how support for each candidate evolves over time as Election Day approaches. Polling data often show non-linear trends, with fluctuations in support that may correspond to campaign events, debates, or shifts in the political landscape. Using natural splines with end_date_num (the number of days since the earliest poll date) allows the model to capture these non-linear trends in a flexible way. By including natural splines with 4 degrees of freedom, we allow the model to capture gradual changes in voter sentiment while avoiding overfitting to noise or abrupt fluctuations in the data. This is particularly important in the context of election polling, where voter opinions can change either slowly or rapidly depending on unpredictable external events. Alternative models, such as standard logistic regression without splines, would fail to capture these changes, potentially leading to a less accurate depiction of how support varies over time.

A unique aspect of our analysis is the focus on battleground states, which have distinct political dynamics that can heavily influence election outcomes. Incorporating state as a random effect provides flexibility in estimating state-specific baseline support levels for each candidate, while still allowing us to identify overall trends across all states. This approach is more effective than treating state as a fixed effect because it avoids overparameterizing the model with individual state coefficients, which could lead to overfitting in cases with limited data for some states. The random effect structure also helps generalize the model’s predictions, making it more robust when predicting support in battleground states with varying polling frequencies.

The Bayesian approach offers advantages for our model by allowing us to incorporate prior information, enhancing the stability and reliability of our estimates. Given that polling data can be sparse for certain states or dates, Bayesian inference helps stabilize parameter estimates, especially when there are limited data points for certain states or specific dates. By using weakly informative priors, we let the data primarily determine the model’s outcomes, while still providing some structure to prevent overfitting or extreme parameter values. The Bayesian framework also produces credible intervals, which provide a clear interpretation of uncertainty around our predictions. In the context of election forecasting, this is valuable as it enables us to quantify the range of plausible outcomes for each candidate’s support, adding a layer of transparency to our predictions.
This model provides interpretability that aligns well with the goals of our paper. By structuring the model with a logit link function, we can interpret the results in terms of odds or probability of victory, which is highly relevant in an election context. This interpretability allows us to understand the likelihood of each candidate winning in a specific state at a given time.

Additionally, the Bayesian Beta regression model offers flexibility in how we examine each candidate’s trends individually. By constructing separate models for Trump and Harris, we can directly compare their temporal trends and state-level variations without confounding effects.

Overall, the Bayesian Beta regression model with a logit link function is the most suitable model for our study. It allows for flexible non-linear trends over time, accounts for regional variation across battleground states, and provides interpretable results in terms of victory probabilities. The Bayesian framework stabilizes estimates in the presence of sparse data and provides meaningful uncertainty estimates. The model’s structure aligns well with our goal of forecasting each candidate’s probability of victory, justifying the choice for achieving a comprehensive analysis and prediction of the 2024 U.S. Presidential Election.

## Model Summary

### Trump Model Summary
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false

@@

# Display the Trump model summary table
knitr::kable(trump_summary) %>%
  kable_styling(full_width = FALSE, position = "center")
```

### Harris Model Summary
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Harris Model."
#| warning: false

@@

@@

# Discussion {#sec-discussion}
## Interpretation of Results {#sec-results-interpretation}
Since the electoral college system makes state-level forecasts crucial, and the U.S. operates under a predominantly two-party system, the focus of election predictions should be on battleground states. While many states are strongly aligned with either the Democratic or Republican party, a handful of key battleground states—where neither party has overwhelming dominance—will likely determine the overall outcome of the election. These battleground states, often evenly split in voter sentiment, hold immense significance because their electoral votes can swing the election in favor of one candidate or the other.

The results of our model underscore the critical role that these battleground states will play in the 2024 election. Although many states are predictably stable due to their party loyalty, it is the six battleground states that will serve as the main arena for competition. In these states, where voter preferences are not as firmly entrenched, even minor shifts in public opinion or turnout could lead to vastly different outcomes. As a result, a comprehensive and precise prediction of voting behavior in these battleground states is far more important than in states where the result is a foregone conclusion.

## Model Performance {#sec-model-performance}
Out-of-sample testing results [Appendix -@sec-model-diagnostics] demonstrate the effectiveness and reliability of our Bayesian logistic regression model in predicting state-level outcomes for the 2024 U.S. presidential election. Our accuracy test generated a 75% score for the model, indicating our model's high reliability for making predictions about state-level election outcomes. Since presidential elections hinge on state-level wins, a reliable state-by-state prediction would imply a reliable national outcome prediction.

>>>>

---
title: "Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory"
subtitle: A Bayesian Analysis of Polling Data from Key Battleground States"
author: 
  - Jimin Lee
  - Sarah Ding
  - Xiyan Chen
thanks: "Code and data are available at: [https://github.com/jamiejiminlee/2024_US_Elections.git](https://github.com/jamiejiminlee/2024_US_Elections.git)."
date: today
date-format: long
abstract: "This paper presents a predictive model for the 2024 United States Presidential Election, focusing on essential battleground states. Utilizing a Bayesian framework and state-level polling data, we estimate the probabilities of victory for candidates Donald Trump and Kamala Harris. Our findings indicate that Harris is projected to secure a win in six out of seven key battleground states, such as Michigan, Nevada, and North Carolina, with predicted support exceeding 50%. This analysis not only forecasts the likely winner of the U.S. election but also equips society and the economy to better undertstand the current U.S. political climate."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction {#sec-intro}
The 2024 United States Presidential Election is anticipated to be a pivotal moment in American political history, particularly as battleground states emerge as key determinants of the electoral outcome. With candidates Donald Trump and Kamala Harris intensifying their campaigns, a thorough understanding of voter preferences in these critical regions becomes essential for forecasting election results. This paper leverages state-level polling data from Project 538 [@FiveThirtyEight] and employs the statistical programming language R [@citeR] to construct a predictive model that estimates the likelihood of victory for each candidate across several crucial battleground states, including Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin.

The primary estimand of this study is the probability of winning for either Donald Trump or Kamala Harris in these battleground states, derived from aggregated polling averages. Our analysis utilizes a Bayesian hierarchical approach combined with natural splines to accurately capture the temporal dynamics of voter sentiment, enabling us to estimate winning probabilities in these decisive areas. By comparing the predicted winning probabilities for both candidates, we can ascertain the likely winner based on prevailing voter support trends. Our findings indicate a competitive landscape, with Kamala Harris projected to secure victories in six of the seven battleground states analyzed, including key states like Michigan and Nevada, where voter sentiment appears to favor her candidacy. However, Donald Trump is also predicted to maintain a competitive edge in Arizona, showcasing the variability in voter preferences across these critical regions. The predicted winner is determined by comparing the estimated probabilities for each candidate, highlighting Harris’s overall advantage leading into the election.

Understanding these predictions is crucial as they not only forecast the likely outcome of the U.S. election but also help society and the economy to better understand the current state of the US political climate. The implications of these findings extend beyond mere predictions, providing a framework for analyzing how shifts in voter sentiment may occur as the election draws near. The remainder of this paper is organized as follows: Section @sec-data details the data and variables utilized in the analysis; Section @sec-model outlines the model setup and estimation strategies employed; Section @sec-results presents the results, including visualizations of the predicted election outcomes; and Section @sec-discussion discusses the implications and limitations of our findings.

# Data {#sec-data}
## Overview
The data for this paper contains the polling information of the 2024 Presidential Election cycle and was obtained from Project 538's polling project [@FiveThirtyEight] on October 19th, 2024. The data is periodically updated throughout the presidential election campaign to reflect current and most up-to-date polling results. It consists of polling information from various pollsters such as the New York Times, ActiVote, Morning Consult, etc. For every poll conducted, relevant variables such as state, start_date, end_date, transparency_score, candidate_name, and percentage of support (pct) for the appropriate candidates are included. The dataset compiles data from all major pollsters in the US and provides us with an understanding of the current state of the presidential election.

The present analysis focuses on Donald Trump and Kamala Harris, the two leading candidates in the current presidential election race. For all the variables mentioned (see @sec-variables), we collected observations for both of the candidates. We collect observations from the dataset where the ‘candidate_name’ is Trump and Harris. We are interested in the support for both candidates over time, for each state, therefore, we collect observations of ‘end_date’, and ‘state’, as well as the electorate support for each candidate is denoted as ‘pct’. Furthermore, since there are many pollsters present in the dataset, we extracted pollsters with high-quality data that is reliable, which is denoted as having a high ‘numeric_grade’ and ‘transparency_score’, details of data cleaning are further explained in [Appendix -@sec-data-details].

## Measurement
Voters’ opinions are multifaceted and can be influenced by numerous factors, including media coverage, recent political events, personal beliefs, and social pressures. Each voter’s decision is shaped by a unique mix of these factors, which can change frequently, especially close to an election. Pollsters start by converting the complex, multi-layered opinions in voters' heads into a simple polling response. This is the first challenge: each voter’s thoughts might not fit neatly into options like “Trump,” “Harris,” or “Undecided.” People may lean toward one candidate but have reservations, or may change their views over time. By asking for a single answer, the poll essentially “forces” an opinion that may not fully capture the voter’s internal conflict.

Once collected, these individual responses are combined into percentages of support. This aggregation abstracts further, treating all “Trump” responses equally, regardless of how confident or reluctant the voter might have been. In effect, the complexity of each opinion is smoothed over into a uniform data point, with many variations of opinions lost in the process.

Finally, these support percentages are fed into a model to make a predictive statement about the election outcome. The model’s goal is to determine who is likely to win, based on the assumption that the aggregated opinions in our data reflect real-world voter behavior on Election Day. However, this step relies heavily on simplification, assuming that the snapshot of opinions captured in the polls will hold steady and that all voters will follow through on their expressed preferences.

The raw polling data gives us percentages of support for each candidate in each state. To measure our outcome variable (the likelihood of winning), we convert these percentages into a binary outcome in our model (predicting Trump or Harris as the winner based on who has higher support in each state).  Our model uses these support percentages to generate a probability of each candidate winning. This transformation helps translate fluctuating polling data into a clear prediction of the likely winner, but it also involves a level of simplification, as it assumes these support percentages accurately reflect voting intentions.

Using state as a predictor variable, which influences our outcome variable, assumes that all responses within a state are comparable and that regional differences are consistent across the state. However, each state has internal diversity—urban vs. rural areas, differences in income levels, or cultural values—that the variable “state” may not fully capture. This simplification is necessary for the model, but it means that nuanced, within-state variations in opinion are lost in the abstraction. Another predictor variable, end_date is also measured, it is designed to capture how voter sentiment changes over time, especially as the election approaches. By using the end date of each poll, the model can recognize patterns or shifts in support due to campaign events, news coverage, or debates. While “end_date” provides temporal context, it simplifies complex, often continuous changes in opinion into a discrete snapshot. This abstraction assumes that voter sentiment is relatively stable between poll dates, which may not reflect reality. For example, an event occurring immediately after a poll’s end date could dramatically shift opinions that the model wouldn’t capture until the next poll. Additionally, the importance of time in influencing voter sentiment is assumed to be similar across states, but certain states might react differently to events depending on their local context or demographic makeup.

## Variables {#sec-variables}
The collected data from Project538 [@FiveThirtyEight] contains several key variables relevant to the analysis:

@@

To ensure the quality of the analysis, only those polls that meet specific thresholds were included - for more details on the data cleaning process, refer to @sec-data-cleaning.

@@

# Put table into kable
knitr::kable(summary_statistics_long)

```
@tbl-outcomevar reveals a total of 107 polls conducted, calculated as the total count of rows in our cleaned dataset. ‘Total polls’ reflects the total number of unique poll entries after filtering for only battleground states, polls with sufficient data quality, and polls that report support percentages for both Trump and Harris. The 107 polls conducted display an average support level of 47.801% for Trump and 47.999% for Harris. This suggests that based on the dataset, Trump has a leading advantage in winning the presidential election race by having a slighty higher average level of support than Harris. The highest percentage of support for Trump in the 107 polls was 51.700% while the lowest recorded support was 43%. For Harris, the highest percentage of support also reached 51.200%, with a minimum of 43.600%. These statistics illustrate the competitive landscape of the presidential election race, without a clear leading candidate in the current race.

@@

#### State
The state ('state') variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of Observations for Each State in Polling Data. Illustrates the number of polling observations across key battleground statesn. Notably, Pennsylvania has the highest count of observations, suggesting a more robust polling presence, while Nevada shows fewer observations. "

@@

#### End Date {#sec-end-date}
End date ('end_date') represents the date the poll ended, essentially the date the poll stopped collecting data from electorates. In the context of our paper, the end date of each poll serves as one of the predictive variables and is used to assess changes in voter support over time. Including end date as a variable in our model can help us identify patterns in voter support as Election Day approaches, potentially reflecting the impact of campaign events, debates, or other political developments. @fig-end-date-distribution displays the distribution of polling end dates from July 2024 to October 2024.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Distribution of Polling End Dates. This histogram displays the frequency of polling end dates for key battleground states, highlighting peaks in polling activity around late September."

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-numgrade-distribution
#| fig-cap: "Density plot illustrating the distribution of numeric grades assigned to pollsters within the polling data. Highlights the frequency of numeric grades, indicating key concentrations that suggest variability in pollster quality. "

@@

#### Transparency Score
Transparency Score ('transparency_score') represents the degree of methodological transparency provided by each pollster, rated on a scale from 0 to 10 by [@FiveThirtyEight]. This score reflects the extent to which pollsters disclose details about their data collection and analysis methods, such as sample size, weighting, and margin of error. In the context of our paper, transparency score is an essential variable for evaluating the reliability of polling data, as higher transparency scores generally indicate a greater level of methodological rigor and openness. @fig-transparency-score-distribution displays the distribution of transparency scores across the polling data for key battleground states. The plot shows that transparency scores cluster heavily around ratings of  7 and 9, suggesting that a majority of polls in our dataset come from sources that maintain a relatively high level of transparency. There are smaller peaks at lower scores (5 and 10), indicating that while some pollsters offer limited transparency, a few polls also achieve perfect transparency scores.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram displaying the distribution of transparency scores among pollsters in the polling data. Indicates the frequency of various transparency scores, highlighting a concentration of scores around 7 and 9, which suggests a predominance of high transparency among the assessed pollsters. "

@@

#### Pollster
Pollster ('pollster') represents the various organizations responsible for conducting polls across key battleground states in the lead-up to the 2024 United States Presidential Election. In the context of our paper, the diversity and frequency of polling by different pollsters are essential for understanding the reliability and variance in the data. Each pollster may use different methodologies, sample populations, and data collection techniques, which can affect the results and introduce variability in polling outcomes.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-pollster-distribution
#| fig-cap: "Horizontal bar chart displaying the distribution of polls conducted by various pollsters in key battleground states. Highlights the frequency of polling activities across different organizations, with Marist and Emerson standing out as the most prolific pollsters."

@@

#### Start Date
Start date ('start_date') represents the date when each poll began collecting responses. In the context of our paper, start date is a key variable as it allows us to focus on polling data collected after July 21, 2024—the date Kamala Harris officially announced her presidential campaign. By filtering the dataset to include only polls with a start date after this announcement, we ensure that each poll reflects voter sentiment with Harris as an active candidate, enabling a more accurate comparison between her and Donald Trump. @fig-start-date-distribution displays the distribution of polling start dates from late July to mid-October 2024, showing when polling efforts began across key battleground states. The histogram reveals peaks in polling activity, particularly in mid-August, early September, and late September, indicating periods of heightened interest or campaign events that likely prompted increased polling.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-start-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling start dates for key battleground states. Displays the frequency of polling activities initiated over time, with a notable peak on September 14, indicating a surge in polling efforts as the election approached."

@@

## Model Justification
The Bayesian Beta Regression model with a logit link function is the most suitable model for predicting voter support for Kamala Harris and Donald Trump in the 2024 U.S. Presidential Election. Our primary outcome variable is the percentage of support for each candidate in seven key battleground states, which is naturally bounded between 0% and 100%. The Beta distribution is ideal for modeling this proportion data, as it allows us to avoid unrealistic predictions outside the (0, 1) range. To capture non-linear trends in voter support as Election Day approaches, we employe natural splines with **end_date_num**, which represents the number of days since the earliest poll date. Additionally, incorporating state-specific random effects acknowledges the variations in voter preferences across different regions, allowing for more accurate modeling of baseline support levels. The Bayesian framework enhances our model by enabling the incorporation of prior information, leading to more stable estimates, especially in the context of sparse polling data. Furthermore, the model's structure allows for clear interpretation in terms of victory probabilities, aligning well with our goals of forecasting electoral outcomes. Further details are discussed in [Appendix -@sec-model-details].

## Model Summary
### Trump Model Summary
@tbl-trump-model-summary fixed effects estimates for several key terms that capture the influence of time on his support in the battleground states. The intercept, with an estimate of -0.0923, represents the baseline log-odds of Trump’s support when other predictors are held constant. This negative intercept suggests a slightly lower baseline level of support. The term nsenddatenumdfEQ42, with an estimate of 0.0803 and a statistically significant positive confidence interval (0.0179 to 0.1426), indicates an increase in support around this time period. Similarly, nsenddatenumdfEQ44 has an estimate of 0.0789 and a confidence interval above zero (0.0014 to 0.1589), also showing a positive effect on support. Other terms such as nsenddatenumdfEQ41 and nsenddatenumdfEQ43 have less significant impacts, suggesting that certain time periods were more influential than others in shaping Trump’s support. The standard errors for the fixed effects terms range from 0.0306 to 0.0676 for nsenddatenumdfEQ43. The relatively low standard error of 0.0306 on the intercept suggests that the baseline estimate is quite stable, meaning we can be more confident about this baseline support level. However, the standard error for nsenddatenumdfEQ43 (0.0676) is higher than for other terms, indicating that this estimate is less precise.
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-model-summary
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false

@@

# Display the Trump model summary table
knitr::kable(trump_summary) %>%
  kable_styling(full_width = FALSE, position = "center")
```
These results imply that support for Trump varied over time in ways that may correspond to specific campaign events or other influences affecting voter sentiment. The significance of terms like nsenddatenumdfEQ42 and nsenddatenumdfEQ44 suggests that changes in voter preference could be captured effectively by this model. And by checking standard errors, helping us determine how much we can rely on those data.

### Harris Model Summary
@tbl-harris-model-summary similarly shows the impact of time on her support in the battleground states. The intercept for the Harris model is estimated at -0.1508, suggesting a slightly lower baseline support level compared to Trump(-0.0923). The term nsenddatenumdfEQ41 has an estimate of 0.1225 and a confidence interval 0.0553 to 0.1903, indicating a significant increase in support during this period. Another positive and significant effect is seen in nsenddatenumdfEQ43, with an estimate of 0.1191 and confidence interval from -0.0071 to 0.2417, which suggests a trend toward increased support. While other terms stay positive, they have less impacts on support. Additionally, the standard errors for fixed effects are generally smaller, ranging from 0.0279 for the intercept to 0.0631 for nsenddatenumdfEQ43. So the larger standard error for nsenddatenumdfEQ43 (0.0631) suggests that there would be some uncertainty in how this time point influences support, though the positive estimate still points toward an upward trend. In comparison to the Trump model, the standard errors in the Harris model are slightly smaller for most terms, indicating potentially more consistent support trends over time in the data analyzed.
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-model-summary
#| tbl-cap: "Fixed Effects Summary for Harris Model."
#| warning: false

@@

```
These results highlight that Harris's support also fluctuated over time, with nsenddatenumdfEQ41 and nsenddatenumdfEQ43 marking periods of notable increases. Helping us understand the pattern, this model predicts Harris's probability of victory and provides insights for future changes in support as the election nears.

@@

# Discussion {#sec-discussion}

## Key Findings {#key-findings}
This paper presents a forecast of the 2024 U.S, Presidential Election, leveraging Bayesian Beta regression models trained on polling data. We were able to model both the electoral support trend over time and the different political dynamics across the battleground states. The model’s predictions reveal that, as Election Day approaches, Kamala Harris holds a lead in six of the seven battleground states–Pennsylvania, Michigan, Wisconsin, Nevada, North Carolina, and Georgia. However, her margins are consistently narrow, with predicted probabilities only slightly above 50% in each of these states. Harris shows minimal leads over Trump, with differences between the probability of winning of no more than 3 percent across all six states. Meanwhile, Trump also holds a narrow advantage in Arizona, suggesting a highly competitive electoral dynamic in this election.

The small margins in the predicted outcomes emphasize the potential volatility of the race; a slight shift in voter sentiment or late-breaking events could easily flip the outcome in these battleground states. This narrow margin of victory suggests a level of electoral fragility for Harris, as her support base may not be solidly locked in. The close race in each state implies that both candidates must maintain their voter engagement and strategic focus, as any shift in these states could have significant implications for the election’s overall outcome. Thus, while Harris is favored in the majority of battleground states, the slight leads indicate a need for cautious interpretation, with both campaigns facing high stakes in these key states.
